{\bf Practical Concurrency Testing}

### 2017-04-24 proposal/talk/proposal.tex
finish writing proposal slides barring some bonuses and picts of mario man

@@

\begin{frame}{Outline}
	guess i should outline my talk huh
\end{frame}

@@

\section{some kinda introductory}

@@

	MCing a transactional program
	\begin{itemize}
		\item Simulate TM, interleaving transactions, rolling-back upon conflicts
			\begin{itemize}
					% TODO: need a bonus slide to justify why this
					% imagine a pair of transactions each with multiple conflicting accesses
					% they should be treated as atomic
				\item slow; invites state space explosion
			\end{itemize}
		\item Enforce atomicity via scheduler, anticipate aborts via analysis
			\begin{itemize}
				\item (under HTM, all transactions might abort)
					% TODO: need a bonus slide on finding when a STM txn may abort.
			\end{itemize}
		\item Suffices to enumerate all {\em observationally-equivalent} schedules.
	\end{itemize}
\end{frame}

@@

@@

>>>>

\begin{frame}{Concurrency!}
	\begin{center}
	\begin{tabular}{ll}
		Initially \texttt{x = 0;} \\
		\\
		{\bf Thread 1} & {\bf Thread 2} \\
		\\
		\texttt{\hilight{orange}{mutex\_lock}(m);}      & \texttt{atomic\_xadd(\&x, 1);} \\
		\texttt{x++;}				   & \texttt{\hilight{olivegreen}{yield}();} \\
		\texttt{\hilight{blue}{mutex\_unlock}(m);}      & \texttt{atomic\_xadd(\&x, 1);} \\
		\texttt{assert(x >= 1);}			& \texttt{assert(x >= 2);} \\
		\\
		\\
		\\
		\\
		\\
	\end{tabular}
	\end{center}
\end{frame}
\begin{frame}{Concurrency!}
	\begin{center}
	\begin{tabular}{ll}
		Initially \texttt{x = 0;} \\
		\\
		{\bf Thread 1} & {\bf Thread 2} \\
		\\
		\texttt{\hilight{orange}{mutex\_lock}(m);} \\
		\texttt{int tmp = x;} \\
								& \texttt{atomic\_xadd(\&x, 1);} \\
								& \texttt{\hilight{olivegreen}{yield}();} \\
								& \texttt{atomic\_xadd(\&x, 1);} \\
		\texttt{x = tmp + 1;} \\
								& \texttt{\hilight{red}{assert(x >= 2);}} \\
		\texttt{\hilight{blue}{mutex\_unlock}(m);} \\
		\texttt{assert(x >= 1);}
	\end{tabular}
	\end{center}
\end{frame}

\begin{frame}{Stateless Model Checking}
	Stateless Model Checking {\em [Godefroid '97]} (MC)
	\begin{itemize}
		\item Test framework controls thread scheduling
		\item Each test iteration, test a different interleaving
		\item Goal: Exhaustively check all possible program behaviours
	\end{itemize}
	\pause
	\linegap

	\begin{center}
		\includegraphics[width=0.25\textwidth]{../../oopsla/exponential-curve.pdf}
	\end{center}
\end{frame}

\begin{frame}{State Space Example}
	\begin{center}
		\includegraphics[width=0.96\textwidth]{../../oopsla/tree-maximal-only.pdf}
	\end{center}
	\linegap

	Possible interleavings represented as a tree
	\begin{itemize}
		\item Node: Intermediate execution state
		\item Edge: State transition from executing a thread
	\end{itemize}
	%\linegap

	%State space size depends on which {\bf preemption points} are used.
\end{frame}

\begin{frame}{Challenges} %\& Existing Coping Techniques}
	Unpredictable state space explosion
	\begin{itemize}
		\item Equivalence reductions: DPOR {\em [Flanagan '05]}, MCR {\em [Huang '15]}
		\item Heuristic search ordering: ICB {\em [Musuvathi '08]}, PCT {\em [Burckhardt '10]}
		\item There will always be a {\bf bigger test case}!
	\end{itemize}
	\pause
	\linegap

	Too much manual intervention for non-expert use
	\begin{itemize}
		\item Manual kernel annotations: Landslide {\em [Blum '12]} {\tiny that's me!}
		\item State space size estimation {\em [Simsa '12]}
		\item Widespread adoption requires {\bf fully automated instrumentation}!
	\end{itemize}
	\pause
	\linegap

	Limited sources of nondeterminism
	\begin{itemize}
		\item Timer-driven schedule nondeterminism (all prior work)
		\item Weak memory nondeterminism {\em [Zhang '15]}
		\item Event-driven applications {\em [Jensen '15]}
		\item {\bf Transactional concurrency}?
	\end{itemize}
\end{frame}

@@

\definecolor{gray}{RGB}{128,128,128}

\begin{frame}{Thesis Statement}
	The Iterative Deepening algorithm makes MC appropriate for both educational and real-world programs.
	\vspace{4.14em}

	Ad-hoc synchronization heuristics make MC accessible to student programmers.
	\vspace{4.14em}

	A new transactional concurrency model keeps MC applicable to the latest real-world techniques.
	\vspace{4.14em}
\end{frame}
\begin{frame}{Outline}
	\hilight{gray}{The Iterative Deepening algorithm makes MC appropriate for both educational and real-world programs.} \\
	\begin{itemize}
		\item {\bf Quicksand and Iterative Deepening}
		\begin{itemize}
			\item Algorithm overview \& data-race preemption points
			\item Verification \& experimental results
		\end{itemize}
	\end{itemize}
	\hilight{gray}{Ad-hoc synchronization heuristics make MC accessible to student programmers.} \\
	\begin{itemize}
		\item {\bf Educational use}
		\begin{itemize}
			\item Pebbles \& 15-410 user study
			\item Pintos \& upcoming plans
		\end{itemize}
	\end{itemize}

	\hilight{gray}{A new transactional concurrency model keeps MC applicable to the latest real-world techniques.} \\
	\begin{itemize}
		\item {\bf Transactional memory}
		\begin{itemize}
			\item Example \& challenges
			\item Equivalence \& failure injection
		\end{itemize}
	\end{itemize}
\end{frame}

%\begin{frame}{Outline}
%	guess i should outline my talk huh
%\end{frame}

@@

\section{Quicksand}

\breakslide{\Large Iterative Deepening of State Spaces}

\begin{frame}{Preemption Points}
	The burning question: ``Which {\bf preemption points} (PPs) are important?''
	\linegap

	State space of interleavings is {\em parameterized} by PPs.
	\begin{itemize}
		\item Too few PPs: Many bugs will go undetected
		\item Preempt everywhere: Completion is often infeasible % say: "for some programs"
	\end{itemize}
	\linegap

	Prior model checkers hard-code a fixed set of PPs, committing to one state space in advance.
	% say: "...then attempt to cope with the resulting exponentially sized problem, however big it may be"
\end{frame}

\begin{frame}{Example (Revisited)}
	\begin{center}
	\begin{tabular}{ll}
		Initially \texttt{x = 0;} \\
		\\
		{\bf Thread 1} & {\bf Thread 2} \\
		\\
		\texttt{\hilight{orange}{mutex\_lock}(m);} \\
		\texttt{int tmp = x;} \\
								& \texttt{atomic\_xadd(\&x, 1);} \\
								& \texttt{\hilight{olivegreen}{yield}();} \\
								& \texttt{atomic\_xadd(\&x, 1);} \\
		\texttt{x = tmp + 1;} \\
								& \texttt{\hilight{red}{assert(x >= 2);}} \\
		\texttt{\hilight{blue}{mutex\_unlock}(m);} \\
		\texttt{assert(x >= 1);}
	\end{tabular}
	\end{center}
\end{frame}

\begin{frame}{State Space (Revisited)}
	\begin{center}
		\includegraphics[width=0.96\textwidth]{../../oopsla/tree-maximal-only.pdf}
		\linegap

		None of these branches contain the necessary preemption!
		\linegap

		A {\bf data-race preemption point} is required to find the bug.
	\end{center}
\end{frame}

\begin{frame}{Iterative Deepening (in Pictures)}
	\begin{center}
	\vspace{-0.88em}
	\includegraphics[width=0.86\textwidth]{../../oopsla/dr-jobs-1.pdf}
	\end{center}
\end{frame}
\begin{frame}{Iterative Deepening (in Pictures)}
	\begin{center}
	\vspace{-0.88em}
	\includegraphics[width=0.86\textwidth]{../../oopsla/dr-jobs-2.pdf}
	\end{center}
\end{frame}
\begin{frame}{Iterative Deepening (in Pictures)}
	\begin{center}
	\vspace{-0.88em}
	\includegraphics[width=0.86\textwidth]{../../oopsla/dr-jobs-3.pdf}
	\end{center}
\end{frame}
\begin{frame}{Iterative Deepening (in Pictures)}
	\begin{center}
	\vspace{-0.88em}
	\includegraphics[width=0.86\textwidth]{../../oopsla/dr-jobs-4.pdf}
	\end{center}
\end{frame}

\begin{frame}{Total Verification}
	\textbf{Convergence theorem:} Testing maximal state space, after detecting all data races, $\equiv$ preempting everywhere!
	\begin{itemize}
		\item Insight: $\forall$ arbitrary buggy interleaving, $\exists$ equivalent interleaving % say 'under DPOR' i guess?
			using only sync/data-race PPs.
	\end{itemize}
	\linegap

	Hence, Iterative Deepening provides ``best of both worlds'':
	\begin{itemize}
		\item Provides total safety guarantee for small tests
		\item Finds bugs quickly for large tests when completion is infeasible
	\end{itemize}
\end{frame}

\begin{frame}{Evaluation}
	79 ``P2'' thread libraries from CMU, 78 ``Pintos'' kernels from U. Chicago \& Stanford
	\linegap

	%629 pairs of \{P2 or Pintos\} + test case
	629 pairs of P2 or Pintos + test case
	\linegap

	Each tested for 10 CPU-hours with
	\begin{itemize}
		\item Quicksand, finding data-race candidates with...
			\begin{itemize}
				\item ...``Pure'' Happens-Before analysis {\em [Lamport '78]}
				\item ...``Limited'' Happens-Before analysis {\em [O'Callahan '03]}
			\end{itemize}
		\item Stand-alone Landslide using ICB {\em [Musuvathi '08]}, preempting at...
			\begin{itemize}
				\item ...synchronization APIs only {\em [Simsa '10]}
				\item ...all shared memory accesses {\em [Holzmann '97]}
			\end{itemize}
	\end{itemize}
\end{frame}

\begin{frame}{Results - Bugs found}
	\begin{center}
		\includegraphics[width=0.8\textwidth]{../../oopsla/bugs-talk.pdf}
	\end{center}
\end{frame}
\begin{frame}{Results - Full verifications}
	\begin{center}
		\includegraphics[width=0.8\textwidth]{../../oopsla/verifs-talk.pdf}
	\end{center}
\end{frame}

\begin{frame}{Deliverables}
	In the thesis, I will:
	\begin{itemize}
		\item Restate the paper {\large \smiley}
		\item Study variance of repeated Quicksand experiment trials
	\end{itemize}
\end{frame}

@@

	MCing a transactional program
	\begin{itemize}
		\item Simulate TM, interleaving transactions, rolling-back upon conflicts
			\begin{itemize}
				\item slow; invites state space explosion
			\end{itemize}
		\item Enforce atomicity via scheduler, anticipate aborts via analysis
			\begin{itemize}
				\item (under HTM, all transactions might abort)
			\end{itemize}
		\item Suffices to enumerate all {\em observationally-equivalent} schedules.
	\end{itemize}
\end{frame}

@@

\breakslide{\Large Bonus Slides}

\begin{frame}{15-410: Summary of bugs found}
	\begin{center}
	\begin{tabular}{r|cc|cc}
	%det bugs found det bugs fixed  races fixed     races found
	& \multicolumn{4}{c}{Number of groups} \\
	Number & \multicolumn{2}{c|}{Deterministic bugs} & \multicolumn{2}{c}{Concurrency bugs} \\
	of bugs & found & fixed & found & fixed \\
	\hline
	0       & 27    & 32    & 15    & 23    \\
	1       & 6     & 5     & 7     & 11    \\
	2       & 9     & 6     & 12    & 8     \\
	3       & 2     & 1     & 7     & 2     \\
	4       & 1     & 1     & 3     & 1     \\
	5       & 2     & 2     & 2     & 1     \\
	11      &       &       & 1     & 1     \\
	\hline
	%Total groups
	%       & 47    & 47    & 47    & 47    \\
	Total bugs
		& 44    & 34    & 85    & 53    \\
	\end{tabular}
	\end{center}
	%\linegap

	%E.g.: 11 groups fixed 1 bug + 8 fixed 2 + 2 $\times$ 3 + 1 $\times$ 4 + 1 $\times$ 5 + 1 overachieving group who fixed 11 = 11 + 16 + 6 + 4 + 5 + 11 = 53.
\end{frame}

@@

\begin{frame}{Interleaving t'xns non-atomically $\rightarrow$ state space explosion}
					% TODO: need a bonus slide to justify why this
					% imagine a pair of transactions each with multiple conflicting accesses
					% they should be treated as atomic
\end{frame}

\begin{frame}{STM: Judging when aborts are possible}
	% TODO
\end{frame}

\breakslide{\Large Super Extra Bonus Slides Copy-Pasted from OOPSLA}

\begin{frame}{Data-Race Definitions}
	\textbf{Data race:} 2 threads access the same memory, and...
	\begin{itemize}
		\item At least one access is a write
		\item Threads do not hold the same lock
		\item No {\em Happens-Before} relation between threads
	\end{itemize}
	\pause
	\linegap

	Variants of Happens-Before (HB)
	\begin{itemize}
		\item {\bf Pure HB:} Any synchronization events {\em [Lamport '78]}
			% Say: "Or another way of putting it I heard from one of yesterday's talks that I'm stealing, one event is guaranteed to be visible to the other"
			% Or don't, idgaf
		\item {\bf Limited HB:} Blocking synchronization (e.g. \texttt{cond\_wait()}) enforces one ordering {\em [O'Callahan '03]}
	\end{itemize}
\end{frame}

\begin{frame}{Happens-Before Example}
	\begin{center}
		\begin{tabular}{ll}
			{\bf Thread 1}  & {\bf Thread 2} \\
			\texttt{\hilight{red}{x++;}}    & \\
			\texttt{mutex\_lock(m);}	& \\
			\texttt{// unrelated}	   & \\
			\texttt{mutex\_unlock(m);}      & \\
				& \texttt{mutex\_lock(m);} \\
				& \texttt{// unrelated} \\
				& \texttt{mutex\_unlock(m);} \\
				& \texttt{\hilight{red}{x++;}} \\
		\end{tabular}
		\linegap

		No race under Pure HB; {\em true potential race} under Limited HB.
	\end{center}
\end{frame}
\begin{frame}{Happens-Before Example}
	\begin{center}
		\begin{tabular}{ll}
			{\bf Thread 1}  & {\bf Thread 2} \\
			\texttt{\hilight{red}{x++;}}    & \\
			\texttt{mutex\_lock(m);}	& \\
			\texttt{\hilight{blue}{y = true;}}      & \\
			\texttt{mutex\_unlock(m);}      & \\
				& \texttt{mutex\_lock(m);} \\
				& \texttt{\hilight{blue}{bool tmp = y;}} \\
				& \texttt{mutex\_unlock(m);} \\
				& \texttt{\hilight{blue}{if (tmp)}~\hilight{red}{x++;}} \\
		\end{tabular}
		\linegap

		No race under Pure HB; {\em false positive} under Limited HB.
	\end{center}
\end{frame}

\begin{frame}{Iterative Deepening Reductions}
	If one job is a subset of another, testing either might let us skip the other.
	\begin{center}
		\includegraphics[width=\textwidth]{../../oopsla/reduction-nothing.png}
	\end{center}
\end{frame}
\begin{frame}{Iterative Deepening Reductions}
	If one job is a subset of another, testing either might let us skip the other.
	\begin{center}
		\includegraphics[width=\textwidth]{../../oopsla/reduction-deferred.png}
	\end{center}
\end{frame}
\begin{frame}{Iterative Deepening Reductions}
	If one job is a subset of another, testing either might let us skip the other.
	\begin{center}
		\includegraphics[width=\textwidth]{../../oopsla/reduction-complete.png}
	\end{center}
\end{frame}

\begin{frame}{Quicksand results - Bugs found by CPU time}
	\begin{center}
		\includegraphics[width=0.8\textwidth]{../../oopsla/bugs-talk.pdf}
	\end{center}
\end{frame}
\begin{frame}{Quicksand results - Bugs found by wall-clock time}
	\begin{center}
		\includegraphics[width=0.8\textwidth]{../../oopsla/bugs-wallclock-talk.pdf}
	\end{center}
\end{frame}
\begin{frame}{Quicksand results - Tests fully verified (CPU time)}
	\begin{center}
		\includegraphics[width=0.8\textwidth]{../../oopsla/verifs-talk.pdf}
	\end{center}
\end{frame}
\begin{frame}{Quicksand results - MC + DR vs single-pass DR}
	\begin{center}
		\includegraphics[width=0.8\textwidth]{../../oopsla/dr-false-negatives-poster.pdf}
	\end{center}
\end{frame}
\begin{frame}{Quicksand results - Representative of real-world programs?}
	Bugs by required number of preemptions to expose:
	\begin{center}
		\footnotesize
	\begin{tabular}{r|c|c}
		{\bf Bound} & {\bf SSS (sync PPs only)} & {\bf SSS (preempt everywhere)} \\
		\hline
		0       & 2     & 1     \\
		1       & 82    & 86    \\
		2       & 16    & 32    \\
		3       & 2     & 3     \\
		4+      & 0     & 0     \\
		\hline
		Total   & 102   & 122   \\
	\end{tabular}
	\end{center}
	\linegap

	Bugs by type (Pure HB trial):
	\begin{itemize}
		\item 56 deadlocks
		\item 49 heap errors (47 use-after-free)
		\item 35 assertion failures
		\item 31 page fault crashes
		\item 1 infinite loop
		\item 1 recursive mutex lock
	\end{itemize}
\end{frame}

\begin{frame}{References}
	\small
	\begin{itemize}
		\item {\bf [Lamport '78]}:
			Leslie Lamport. Time, clocks, and the ordering of events in a distributed system.
			Commun. ACM, 1978.
		\item {\bf [Godefroid '97]}: Patrice Godefroid.
			VeriSoft. A tool for the automatic analysis of concurrent reactive software. CAV 1997.
		\item {\bf [Holzmann '97]}: The model checker SPIN. TOSE 1997.
		\item {\bf [Engler '03]}: Dawson Engler and Ken Ashcraft.
			RacerX: effective, static detection of race conditions and deadlocks. SOSP 2003.
		\item {\bf [O'Callahan '03]}: Robert O'Callahan and Jong-Deok Choi.
			Hybrid dynamic data race detection. PPoPP '03.
		\item {\bf [Flanagan '05]}: Cormac Flanagan and Patrice Godefroid. Dynamic partial-order reduction for
			model checking software. POPL 2005.
		\item {\bf [Musuvathi '08]}: Madanlal Musuvathi et al. Finding and Reproducing Heisenbugs in Concurrent
			Programs. OSDI 2008.
	\end{itemize}
\end{frame}
\begin{frame}{References}
	\small
	\begin{itemize}
		\item {\bf [Buckhardt '10]}: A randomized scheduler with probabilistic guarantees of finding bugs. ASPLOS 2010.
		\item {\bf [Simsa '12]}: Jiri Simsa. Runtime Estimation and Resource Allocation for
			Concurrency Testing. CMU-PDL-12-113. December 2012.
		\item {\bf [Blum '12]}: Ben Blum. Landslide: Systematic Dynamic Race Detection in Kernel
			Space. CMU-CS-12-118. May 2012.
		\item {\bf [Kasicki '12]}: Baris Kasicki. Data races vs. data race bugs: telling the difference with Portend. ASPLOS '12.
		\item {\bf [Huang '15]}: Jeff Huang. Stateless model checking concurrent programs with maximal causality reduction. PLDI '15.
		\item {\bf [Jensen '15]}: Stateless Model Checking of Event-Driven Applications. OOPSLA 2015.
	\end{itemize}
\end{frame}
